Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2018-07-13 paper.tex
(>) 4 spaces indentation
...

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers) and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot use long-lasting locals and loops, which are
elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \emph{goto}''.%
%\footnote{``Callbacks as our Generations' \emph{goto} Statement'':
%\url{tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

In this work, we advocate structured synchronous reactive programming (SSRP) as
a more productive alternative for game logic development.
We present a qualitative case study of rewriting Pingus from C++ to \CEU.
%
\CEU~\cite{ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
programming language that originally targets embedded soft real-time systems.
It aims to offer a concurrent, safe, and expressive alternative to C.
%
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
%\CEU provides primitives that help describing complex control-flow
%relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage.
The runtime is single threaded and does not rely on garbage collection for
memory management~\cite{ceu.sensys13}.
%
Existing work in the context of embedded sensor networks evaluates the
expressiveness of \CEU in comparison to event-driven code in C and attests a
reduction in source code size (around 25\%) with a small increase in memory
usage (around 5--10\%)~\cite{ceu.sensys13}.
%
\CEU has also been used in the context of multimedia
systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}.

Our case study shows gains in productivity for four selected behaviors in the
game logic of Pingus rewritten in \CEU.
We present an in-depth qualitative analysis of the proposed solutions in
comparison to the original implementations in C++.
%
Not all techniques result in reduction of \locs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
%
We also identify four control-flow patterns that likely apply to most games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
%
A control-flow pattern is a recurring technique to describe execution
dependency and/or explicit ordering between statements.
% (or groups of statements).
%For instance, consider how a key press stimulus propagates through the game
%entities and also what happens with them if the stimulus causes the end of the
%game.
%We focus on a qualitative analysis for the programming techniques that we
%applied during the rewriting process.

We employed a \emph{live code rewrite}, i.e., starting from the original
codebase in C++, we reimplemented it piece-by-piece in \CEU without breaking
the game compilation and execution.
%
This approach shows the feasibility of a partial and gradual translation
between the languages.

@@

\CEU is a synchronous reactive language in which programs evolve in a sequence
of discrete reactions to external events with the characteristics that follow:
%
\begin{itemize}
\item \textbf{Reactive:} code only executes in reactions to events.
\item \textbf{Structured:} programs use structured control mechanisms, such as
    \code{spawn} and \code{await} (to create and suspend activities).
\item \textbf{Synchronous:} reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{itemize}
%
\CEU is designed for control-intensive applications, supporting concurrent
lines of execution, known as \emph{trails}, and instantaneous broadcast
communication through events.
%
Computations within a reaction (such as expressions, assignments, and system
calls) are also instantaneous considering the synchronous
hypothesis~\cite{rp.hypothesis}.
%
%\CEU provides an \code{await} statement which is the only that actually
%``consumes'' time.
%
\CEU provides an \code{await} statement that blocks the current running trail
allowing the program to execute its other trails; when all trails are blocked,
the reaction terminates and control returns to the environment.

@@

Listing~\ref{lst.syntax} shows a compact reference of \CEU and
Listing~\ref{lst.ceu} shows a simple example that blinks an LED and terminates
on a button press or after 1 hour of execution.
%
The example declares two external events (ln~1--2): an input that represents
a button and an output that represents an LED.
%
The blink behavior uses a \code{code} abstraction (ln~4--11) which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}, but which can also
interact directly with the environment through \code{await} statements
(ln.~7,9).
%
The \code{spawn} statement (ln~13) instantiates a \code{Blink} abstraction to
execute in the background and the program then creates two trails to await
two events at the same time (ln~14--18).
Since the composition is a \code{par/or}, it will terminate when either of the
two events occur proceeding to the end of the file and consequently also
terminating the program.

In \CEU, every execution path within loops must contain at least one
\code{await} statement to an external input
event~\cite{ceu.sensys13,esterel.primer}.
%
On the one hand, \CEU shares a restriction with event-driven programming:
computations that take a non-negligible time to run (e.g., cryptography or
image processing algorithms) violate the zero-delay hypothesis, and thus cannot
be directly implemented.
%
On the other hand, all reactions to the external environment are guaranteed to
be computed in bounded time, ensuring that games progress with
time~\cite{ceu.lctes18}.

@@

\begin{figure*}
{\footnotesize
\begin{verbatim}
          Path             Ceu     C++   Ceu/C++    Description
          ------------    ----    ----    ----      --------------------------------------
     1    game/           2064    2268    0.91      the main gameplay
     2      ./             710     679    1.05        main functionality
     3      objs/          470     478    0.98        world objects (tiles, traps, etc)
     4      pingu/         884    1111    0.80        pingu behaviors
     5        ./           343     458    0.75          main functionality
     6        actions/     541     653    0.83          pingu actions (bomber, climber, etc)
     7    worldmap/        468     493    0.95      campaign worldmap
     8    screens/        1109    1328    0.84      menus and screens
     9        option/      347     357    0.97        option menu
    10        others/      762     971    0.78        other menus and screens
    11    misc/             56      46    1.22      miscellaneous functionality
                          ----    ----    ----
                          3697    4135    0.89
\end{verbatim}
}
\caption{The Pingus codebase directory tree.
\label{tab.tree}
}
\end{figure*}

@@

Although the analysis in this work is qualitative, the rows with lower ratio
numbers in Figure~\ref{tab.tree} do correlate with the parts of the game
logic that we consider more susceptible to structured reactive programming.
For instance, the \emph{Pingu} behavior (row 4, \emph{ratio 0.80}) contains
complex animations that are affected by timers, game rules, and user
interaction.
In contrast, the \emph{Option screen} (row 9, \emph{ratio 0.97}) is a simple UI
grid with trivial mouse interactions.

@@

@@

Listing~\ref{lst.armageddon.a} shows the C++ implementation for the class
\code{ArmageddonButton} with methods for rendering the button and handling
mouse and timer events.
The code in the figure focus on the double click detection and hides unrelated
parts with \code{<...>}.
%
The methods \code{update} (ln 14--26) and \code{on\_click} (ln 28--34) are
examples of \emph{short-lived callbacks}, which are pieces of code that execute
atomically in reaction to external input events.
The callback \code{on\_click} reacts to mouse clicks detected by the base class
\code{RectComponent} (ln 2), while the callback \code{update} continuously
reacts to the passage of time, frame by frame.
%Callbacks are short lived because they must react to input as fast as possible
%to let other callbacks execute, keeping the game with real-time responsiveness.
%
The class first initializes the variable \code{pressed} (ln 3) to track the
first click (ln 32).
It also initializes the variable \code{press\_time} (ln 4) to count the time
since the first click (ln 16--17).
If another click occurs within 1 second, the class signals the double click to
the application (ln 29--30).
Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
(ln 18--21).
%
Figure~\ref{fig.armageddon.fsm} illustrates how we can model the double-click 
behavior in C++ as a state machine.
The circles represent the state of the variable \code{pressed}, and the arrows 
represent the callbacks manipulating it.
%
Note in Listing~\ref{lst.armageddon.a} how the accesses to the state variables
are spread across the entire class: the distance between the initialization of
\code{pressed} (ln  3) and the last access to it (ln 32) is over 40 lines in
the original file.
Arguably, this dispersion of code across methods makes the understanding and 
maintenance of the double-click behavior more difficult.
Also, even though the state variables are private, unrelated methods such as 
\code{draw}, which is defined in middle of the class (ln 10--12), can
potentially access them.

\CEU supports structured constructs to deal with events, aiming to eradicate
explicit manipulation of state variables for control-flow purposes.
%
In Listing~\ref{lst.armageddon.b}, the loop to detect double clicks (ln 4--12)
awaits the first click (ln 5) and then, awaits either a 1-second timeout (ln 7)
or the second click (ln 9).
If the second click occurs within 1 second, the \code{break} terminates the
loop (ln 10) and the \code{emit} in sequence signals the double click to the
application (ln 14).
Otherwise, the \code{par/or} block as a whole aborts after 1 second  and the
loop restarts, falling back to the first click \code{await} (ln 5).
%
Double click detection in \CEU does not rely on state variables and is entirely
self-contained in the \code{loop} body.
Also, those 7 lines of code \emph{only} detect the double click, leaving the
actual effect (ln 12) as well as all unrelated code (such as redrawing the
button) to happen outside the loop.

@@

Pingus supports 16 actions in the game.
Five of them implement at least one state machine and are considerable smaller
in \CEU in terms of \locs (Figure~\ref{tab.actions}).
%(after eliminating the state variables).
%
For the other 11 actions without state machines, the reduction in \locs is
negligible.
%
This asymmetry illustrates the gains in expressiveness when describing state
machines in direct style.

@@

Among all 65 implementation files in \CEU, we found 29 cases in 25 files that
use structured mechanisms to substitute states machines.
They typically manifest as \code{await} statements in sequence (e.g.,
ln 5,7 in Listing~\ref{lst.armageddon.b}).
%or in aborting constructs such as \code{par/or} and \code{watching}.

@@

\subsubsection{ Transition from \emph{Story}  to \emph{Credits} Screen}

\begin{lstlisting}[
  numbers=left,xleftmargin=0.5cm,
  float=t,
  caption={C++: Transition from \emph{Story} to \emph{Credits screen}.},
  label={lst.story.a},
]
StoryDot::StoryDot(FileReader& reader) :
    show_credits(false), // do not show by default
{
    <...>
    reader.read("credits", show_credits);
}             // from file

@@

\begin{lstlisting}[
  numbers=left,xleftmargin=0.5cm,
  float=t,
  caption={\CEU: Transition from \emph{Story} to \emph{Credits screen}.},
  label={lst.story.b},
]
loop do
    var int ret = await Worldmap();
    if ret==CREDITS or ret==BACK then
      <...>
      var bool is_click = await Story();
      if is_click and ret==CREDITS then
        <...>
        await Credits();
      end
    else
      <...>
    end
end
\end{lstlisting}

The campaign world map has clickable blue dots in the two extremes of the
map road to show introductory and closing ambience stories, respectively.
For introductory stories, the game returns to the world map after showing 
the story pages.
For closing stories, the game also shows a \emph{Credits screen} before
returning to the world map.%
\footnote{Credits screen animation: \url{github.com/an000/p/#4} }

In \CEU, the \code{loop} of Listing~\ref{lst.story.b} controls the flow between
the screens to show as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln 2), which shows the map and lets
the player interact with it (e.g., walking around) until a dot is clicked.
If the player selects a story dot (ln 4--9), we invoke the \code{Story}
and await its termination (ln 5).
After showing the story, we check the returned values (ln 6) to perhaps show
the \code{Credits} screen (ln 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

@@

Continuation passing typically controls the overall structure of the game in
C++, such as screen transitions in menus and level progressions.
%
\CEU adopts the direct style technique in five cases involving screen
transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches between pingu
actions during gameplay (e.g., \emph{walking} to \emph{falling} and back to
\emph{walking}).

@@

@@

In C++, the class \code{Bomber} in Listing~\ref{lst.hier.a} declares a
\code{sprite} member (ln 3) to handle its animation frames.
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus to the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{ScreenManager::display} in the main game loop calls\\
      \code{ScreenManager::update} when starting a new frame.
\item \code{ScreenManager::update} calls \code{screen->update} for the active
      game screen (i.e., a \code{GameSession} instance, considering the screen
      in which the \code{Bomber} appears).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{objs->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is indeed necessary considering the game
architecture:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item With a single assignment to \code{screen}, one can easily deactivate the
current screen and redirect all dispatches to a new screen (step 2).
\item The \code{World} class manages and dispatches events to all game
      entities with a common interface \code{WorldObj}, such as the pingus and
      traps (step 4).
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus (step 5).
\item Since a single pingu can change its actions during lifetime, the
      \code{action} member decouples them with another level of indirection
      (step 6).
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions (step 7).
\end{itemize}
%
Like \code{update}, the \code{draw} callback also flows through a similar
dispatching hierarchy until reaching the \code{Sprite} class.

In \CEU, the \code{Bomber} abstraction presented in Listing~\ref{lst.hier.b}
spawns a \code{Sprite} animation instance on its body (ln 3).
%
The \code{Sprite} abstraction can react directly to external \code{update}
and \code{draw} events, bypassing the program hierarchy entirely.
Events in \CEU are broadcasted to the entire application in lexical order,
i.e., an abstraction that appears first in the source code (e.g., ln 3) reacts
before another one that appears second (e.g., hidden in ln 4).
%As discussed in Section~\ref{sec.pats.fsms.2}
This rule preserves determinism and also conforms to the program static
hierarchy.
While (\emph{and only while}) the bomber abstraction is alive, the sprite
animation remains alive and reacts to the \code{update} and \code{draw} events.
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.
% TODO: hidden in the language runtime
%Note that one can still declare a local event to restrict its visibility like
%a local variable.

@@

Note that dispatching hierarchies cross game engine code, suggesting that most
games also rely heavily on this control-flow pattern.
In the case of the Pingus engine, we rewrote 9 files with a reduction from 515
to 173 \locs (not listed in Figure~\ref{tab.tree}), mostly due to dispatching
code removal.

@@

A number of domain-specific languages, frameworks, and techniques have been
proposed for particular subsystems of the game logic, such as
animations~\cite{games.anims.2006,games.anims.2003,games.anims.1996,games.anims.1982},
game state and screen progression~\cite{games.fsms.2006.1,games.fsms.2006.2}, and
behavior and AI modeling~\cite{games.bts,games.bts.unreal}.
%
In Pingus, the adoption of \CEU is not restricted to a specific subsystem.
We employed \CEU at the very core of the game for event dispatching
(Section~\ref{sec.pats.dispatching}) and memory management of entities
(Section~\ref{sec.pats.lifespan}), eliminating parts of the original game
engine.
We also implemented all entity animations and behaviors
(Section~\ref{sec.pats.fsms}), and screen transitions
(Section~\ref{sec.pats.cps})
%, and intermodule communication (Section~\ref{sec.pats.signaling})
using the available control mechanisms of \CEU.
%
Furthermore, \CEU is a superset of C targeting reactive systems in general, not
only games, and has also been successfully adopted in other domains, such as
    wireless sensor networks~\cite{ceu.sensys13,ceu.terra} and
    multimedia systems~\cite{ceumedia.webmedia16}.

@@

Overall, we most difficulties in implementing control-flow behavior in game
logic is not inherent to this domain, but a result of accidental complexity due
to the lack of structured abstractions and an appropriate concurrency model to
develop event-based applications.

>>>>

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers) and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot use long-lasting locals and loops, which are
elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \emph{goto}''.%
\footnote{``Callbacks as our Generations' \emph{goto} Statement'':
\url{tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

In this work, we advocate structured synchronous reactive programming (SSRP) as
a more productive alternative for game logic development.
We present a qualitative case study of rewriting Pingus from C++ to \CEU.
%
\CEU~\cite{ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
programming language that originally targets embedded soft real-time systems.
It aims to offer a concurrent, safe, and expressive alternative to C.
%
SSRP lets developers write code in direct style, recovering from the inversion
of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
%\CEU provides primitives that help describing complex control-flow
%relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage.
The runtime is single threaded and does not rely on garbage collection for
memory management~\cite{ceu.sensys13}.
%
Existing work in the context of embedded sensor networks evaluates the
expressiveness of \CEU in comparison to event-driven code in C and attests a
reduction in source code size (around 25\%) with a small increase in memory
usage (around 5--10\%) and comparable CPU responsiveness~\cite{ceu.sensys13}.
%
\CEU has also been used in the context of multimedia
systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}.

Our case study shows gains in productivity for four selected behaviors in the
game logic of Pingus rewritten in \CEU.
We present an in-depth qualitative analysis of the proposed solutions in
comparison to the original implementations in C++.
%
Not all techniques result in reduction of \locs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.
%
We also identify four control-flow patterns that likely apply to most games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
%
In this context, a control-flow pattern is a recurring technique to describe
execution dependency and/or explicit ordering between statements.
% (or groups of statements).
%For instance, consider how a key press stimulus propagates through the game
%entities and also what happens with them if the stimulus causes the end of the
%game.
%We focus on a qualitative analysis for the programming techniques that we
%applied during the rewriting process.

We employed a \emph{live code rewrite}, i.e., starting from the original
codebase in C++, we reimplemented it piece-by-piece in \CEU without breaking
the game compilation and execution.
%
This approach shows the feasibility of a partial and gradual translation
between the languages, and also the possibility of keeping performance-critical
functionality in C++, such as in heavy graphical primitives.

@@

@@

\CEU is a synchronous reactive language in which programs evolve in a sequence
of discrete reactions to external events with the characteristics that follow:
%
\begin{itemize}
\item \textbf{Reactive:} code only executes in reactions to events.
\item \textbf{Structured:} programs use structured control mechanisms, such as
    \code{spawn} and \code{await} (to create and suspend activities).
\item \textbf{Synchronous:} reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{itemize}
%
\CEU is designed for control-intensive applications, supporting concurrent
lines of execution, known as \emph{trails}, and instantaneous broadcast
communication through events.
%
Computations within a reaction (such as expressions, assignments, and system
calls) are also instantaneous considering the synchronous
hypothesis~\cite{rp.hypothesis}.
%
%\CEU provides an \code{await} statement which is the only that actually
%``consumes'' time.
%
\CEU provides an \code{await} statement that blocks the current running trail
allowing the program to execute its other trails; when all trails are blocked,
the reaction terminates and control returns to the environment to process
upcoming events (e.g., mouse clicks and timers).

Listing~\ref{lst.syntax} shows a compact reference of \CEU and
Listing~\ref{lst.ceu} shows a simple example that blinks an LED and terminates
either on a button press or after 1 hour of execution.
%
The example declares two external events (ln~1--2): an input that represents
a button and an output that represents an LED.
%
The blink behavior uses a \code{code} abstraction (ln~4--11) which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}, but which can also
interact directly with the environment through \code{await} statements
(ln.~7,9).
%
The \code{spawn} statement (ln~13) instantiates a \code{Blink} abstraction to
execute in the background and the program then creates two trails to await
two events at the same time (ln~14--18).
Since the composition is a \code{par/or}, it will terminate when either of the
two events occur proceeding to the end of the file and consequently also
terminating the program.

In \CEU, every execution path within loops must contain at least one
\code{await} statement to an external input
event~\cite{ceu.sensys13,esterel.primer}.
%
On the one hand, \CEU shares a restriction with standard event-driven
programming: computations that take a non-negligible time to run (e.g.,
cryptography or image processing algorithms) violate the zero-delay hypothesis,
and thus cannot be directly implemented.
%
On the other hand, all reactions to the external environment are guaranteed to
be computed in bounded time~\cite{ceu.lctes18}, ensuring that games progress
with time.

@@

\begin{figure*}
{\footnotesize
\begin{verbatim}
          Path             Ceu     C++   Ceu/C++    Description
          ------------    ----    ----    ----      ----------------------------------------
     1    game/           2064    2268    0.91      the main gameplay
     2      ./             710     679    1.05        main functionality
     3      objs/          470     478    0.98        world objects (tiles, traps, etc)
     4      pingu/         884    1111    0.80        pingu behaviors
     5        ./           343     458    0.75          main functionality
     6        actions/     541     653    0.83          pingu actions (bomber, climber, etc)
     7    worldmap/        468     493    0.95      campaign worldmap
     8    screens/        1109    1328    0.84      menus and screens
     9      option/        347     357    0.97        option menu
    10      others/        762     971    0.78        other menus and screens
    11    misc/             56      46    1.22      miscellaneous functionality
                          ----    ----    ----
                          3697    4135    0.89
\end{verbatim}
}
\caption{The Pingus codebase directory tree.
\label{tab.tree}
}
\end{figure*}

@@

Although the analysis in this work is qualitative, the rows with lower ratio
numbers in Figure~\ref{tab.tree} do correlate with the parts of the game
logic that we consider more susceptible to SSRP.
For instance, the \emph{Pingu} behavior (row 4, \emph{ratio 0.80}) contains
complex animations that are affected by timers, game rules, and user
interaction.
In contrast, the \emph{Option screen} (row 9, \emph{ratio 0.97}) is a simple UI
grid with trivial mouse interactions.

@@

Listing~\ref{lst.armageddon.a} shows the C++ implementation for the class
\code{ArmageddonButton} with methods for rendering the button and handling
mouse and timer events.
The code in the figure focus on the double click detection and hides unrelated
parts with \code{<...>}.
%
The methods \code{update} (ln 14--26) and \code{on\_click} (ln 28--34) are
examples of \emph{short-lived callbacks}, which are pieces of code that execute
atomically in reaction to external input events.
The callback \code{on\_click} reacts to mouse clicks detected by the base class
\code{RectComponent} (ln 2), while the callback \code{update} continuously
reacts to the passage of time, frame by frame.
%
%Callbacks are short lived because they must react to input as fast as possible
%to let other callbacks execute, keeping the game with real-time responsiveness.
%
The class first initializes the variable \code{pressed} (ln 3) to track the
first click (ln 32).
It also initializes the variable \code{press\_time} (ln 4) to count the time
since the first click (ln 16--17).
If another click occurs within 1 second, the class signals the double click to
the application (ln 29--30).
Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
(ln 18--21).
%
Figure~\ref{fig.armageddon.fsm} illustrates how we can model the double-click 
behavior in C++ as a state machine.
The circles represent the state of the variable \code{pressed}, and the arrows 
represent the callbacks manipulating it.
%
Note in Listing~\ref{lst.armageddon.a} how the accesses to the state variables
are spread across the entire class: the distance between the initialization of
\code{pressed} (ln  3) and the last access to it (ln 32) is over 40 lines in
the original file.
Arguably, this dispersion of code across methods makes the understanding and 
maintenance of the double-click behavior more difficult.
Also, even though the state variables are private, unrelated methods such as 
\code{draw}, which is defined in middle of the class (ln 10--12), can
potentially access them.

@@

\CEU supports structured constructs to deal with events, aiming to eradicate
explicit manipulation of state variables for control-flow purposes.
%
In Listing~\ref{lst.armageddon.b}, the loop to detect double clicks (ln 4--12)
awaits the first click (ln 5) and then, awaits either a 1-second timeout (ln 7)
or the second click (ln 9).
If the second click occurs within 1 second, the \code{break} terminates the
loop (ln 10) and the \code{emit} in sequence signals the double click to the
application (ln 14).
Otherwise, the \code{par/or} block as a whole aborts after 1 second  and the
loop restarts, falling back to the first click \code{await} (ln 5).
%
Double click detection in \CEU does not rely on state variables and is entirely
self-contained in the \code{loop} body.
Also, those 9 lines of code \emph{only} detect the double click, leaving the
actual effect (ln 14) as well as all unrelated code (such as redrawing the
button) to happen outside the loop.

@@

In Pingus, the player may assign actions to specific pingus, as illustrated in
Figure~\ref{fig.bomber.action}.
%
The \emph{Bomber action} explodes the clicked pingu, throwing particles around
and also destroying the terrain under its radius.%
\footnote{Bomber action animation: \url{github.com/an000/p/#2} }
%
We model the explosion animation with a sequential state machine with effects
associated to specific frames, such as playing a sound and throwing the
particles.
%
Pingus supports other 15 actions in the game.
Five of them implement at least one state machine and are considerable smaller
in \CEU in terms of \locs (Figure~\ref{tab.actions}).
%(after eliminating the state variables).
%
For the other 11 actions without state machines, the reduction in \locs is
negligible.
%
This asymmetry illustrates the gains in expressiveness when describing state
machines in direct style.

@@

Among all 65 implementation files in \CEU, we found 29 cases in 25 files that
use structured mechanisms to substitute states machines.
They typically manifest as \code{await} statements in sequence (e.g.,
ln 5,9 in Listing~\ref{lst.armageddon.b}).
%or in aborting constructs such as \code{par/or} and \code{watching}.

@@

\subsubsection{ Transition from \emph{Story}  to \emph{Credits} and \emph{Worldmap} Screen}

The campaign world map has clickable blue dots in the two extremes of the
map road to show introductory and closing ambience stories, respectively.
For introductory stories, the game returns to the world map after showing 
the story pages.
For closing stories, the game also shows a \emph{Credits screen} before
returning to the world map.%
\footnote{Credits screen animation: \url{github.com/an000/p/#4} }
From the click in the story dot until the return to the world map, the game
animates the story with a timer-based scrolling and also reacts to user input
to advance it.

\begin{lstlisting}[
  numbers=left,xleftmargin=0.5cm,
  float=t,
  caption={C++: Transition from \emph{Story} to \emph{Credits} and \emph{Worldmap} screen.},
  label={lst.story.a},
]
StoryDot::StoryDot(FileReader& reader) :
    show_credits(false), // do not show by default
{
    <...>
    reader.read("credits", show_credits);
}             // from file

@@

\begin{lstlisting}[
  numbers=left,xleftmargin=0.5cm,
  float=t,
  caption={\CEU: Transition from \emph{Story} to \emph{Credits} and \emph{Worldmap} screen.},
  label={lst.story.b},
]
loop do
    var int ret = await Worldmap();
    if ret==CREDITS or ret==BACK then
        <...>
        var bool is_click = await Story();
        if is_click and ret==CREDITS then
            <...>
            await Credits();
        end
    else
        <...>
    end
end
\end{lstlisting}

In \CEU, the \code{loop} of Listing~\ref{lst.story.b} controls the flow between
the screens as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln 2), which shows the map and lets
the player interact with it (e.g., walking around) until a dot is clicked.
If the player selects a story dot (ln 4--9), we invoke the \code{Story}
and await its termination (ln 5).
After showing the story, we check the returned values (ln 6) to perhaps show
the \code{Credits} screen (ln 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

@@

Continuation passing typically controls the overall structure of games in, such
as screen transitions in menus and level progressions.
%
In Pingus, \CEU adopts the direct style technique in five cases involving
screen transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches between pingu
actions during gameplay (e.g., \emph{walking} to \emph{falling} and back to
\emph{walking}).

@@

In C++, the class \code{Bomber} in Listing~\ref{lst.hier.a} declares a
\code{sprite} member (ln 3) to handle its animation frames.
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus to the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{ScreenManager::display} in the main game loop calls\\
      \code{ScreenManager::update} when starting a new frame.
\item \code{ScreenManager::update} calls \code{screen->update} for the active
      game screen (i.e., a \code{GameSession} instance, considering the screen
      in which the \code{Bomber} appears).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{objs->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is indeed necessary considering the typical
OO game architecture:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item With a single assignment to \code{screen}, one can easily deactivate the
current screen and redirect all dispatches to a new screen (step 2).
\item The \code{World} class manages and dispatches events to all game
      entities with a common interface \code{WorldObj}, such as the pingus and
      traps (step 4).
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus (step 5).
\item Since a single pingu can change its actions during lifetime, the
      \code{action} member decouples them with another level of indirection
      (step 6).
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions (step 7).
\end{itemize}
%
Like \code{update}, the \code{draw} callback also flows through a similar
dispatching hierarchy until reaching the \code{Sprite} class.

@@

In \CEU, the \code{Bomber} abstraction presented in Listing~\ref{lst.hier.b}
spawns a \code{Sprite} animation instance on its body (ln 3).
%
However, in \CEU, the \code{Sprite} abstraction can react directly to external
\code{update} and \code{draw} events, bypassing the program hierarchy entirely.
Events in \CEU are broadcasted to the entire application in lexical order,
i.e., an abstraction that appears first in the source code (e.g., ln 3) reacts
before another one that appears second (e.g., hidden in ln 4).
%As discussed in Section~\ref{sec.pats.fsms.2}
This rule preserves determinism and also conforms to the program static
hierarchy.
While (\emph{and only while}) the bomber abstraction is alive, the sprite
animation remains alive and reacts to the \code{update} and \code{draw} events.
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.
% TODO: hidden in the language runtime
%Note that one can still declare a local event to restrict its visibility like
%a local variable.

@@

Note that dispatching hierarchies cross game engine code, suggesting that most
games also rely heavily on this control-flow pattern.
In the case of the Pingus engine, we rewrote 9 of its files with a reduction
from 515 to 173 \locs, mostly due to dispatching code removal (not listed in
Figure~\ref{tab.tree}, since it's engine code).

@@

A number of domain-specific languages, frameworks, and techniques have been
proposed for particular subsystems of the game logic, such as
animations~\cite{games.anims.2006,games.anims.2003,games.anims.1996}, %,games.anims.1982
game state and screen progression~\cite{games.fsms.2006.1,games.fsms.2006.2}, and
behavior and AI modeling~\cite{games.bts,games.bts.unreal}.
%
In Pingus, the adoption of \CEU is not restricted to a specific subsystem.
We employed \CEU at the very core of the game for event dispatching
(Section~\ref{sec.pats.dispatching}) and memory management of entities
(Section~\ref{sec.pats.lifespan}), eliminating parts of the original game
engine.
We also implemented all entity animations and behaviors
(Section~\ref{sec.pats.fsms}), and screen transitions
(Section~\ref{sec.pats.cps})
%, and intermodule communication (Section~\ref{sec.pats.signaling})
using the available control mechanisms of \CEU.
%
Furthermore, \CEU is a superset of C targeting reactive systems in general, not
only games, and has also been successfully adopted in other domains, such as
    wireless sensor networks~\cite{ceu.sensys13,ceu.terra} and
    multimedia systems~\cite{ceumedia.webmedia16}.

@@

Overall, we consider that most difficulties in implementing control-flow
behavior in game logic is not inherent to this domain, but a result of
accidental complexity due to the lack of structured abstractions and an
appropriate concurrency model to develop event-based applications.
